Binning Strategies and Related Loss for Large Data

### 2013-03-31 Paper.Rnw
Edits made to plots and figures so address issues we discussed
Also, some code chunks are effectively commented out using "eval=FALSE"
because their code is not running properly and may not be used in
revised version, if needed they will be reworked.

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, echo=F,include=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw()
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw()
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw()
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() +
  theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

@@

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")
@

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints2.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Loss in Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE>>=
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
    qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
}

@@

\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
\noindent
Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.

<<tablelossexample, echo=FALSE, results='asis',cache=FALSE>>=
loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)

@@

table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")

>>>>

  res <- data.frame(sapply(1:(k-1), function(var) {
			return(rep(x[,var], x[,k]))
		}
	))
	names(res) <- names(x)[1:(k-1)]
	return(res)
}
d1.ext <- xtd.data(d1)
@
<<scatterplots, echo=F,include=F>>=
qplot(data=d1.ext, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", alpha=I(.05), size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingAlpha.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", size=I(3)) + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/Overplotting.pdf", width=5, height=5)
qplot(data=d1, x=G, y=SO, geom="point", main="Games and Strike Outs in ML Baseball", shape="o", size=I(6)) + scale_shape_identity() + theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)")
  ggsave("./figure/OverplottingCircles.pdf", width=5, height=5)
qplot(data=d1.ext, x=G, y=SO, geom="hex", fill=..count.., group=1) +  
  scale_fill_gradient("Frequency", low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 10, 100, 1000)) + theme_bw() + xlab("Games (Count)") + ylab("Strike Outs (Count)") + theme(legend.position="bottom")
  ggsave("./figure/HexBinning.pdf", width=5, height=5)
d1.hex.obj <- hexbin(d1.ext$G, d1.ext$SO)
d1.hex <- data.frame(hcell2xy(d1.hex.obj))
names(d1.hex) <- c("G", "SO")
d1.hex$Freq <- d1.hex.obj@count
qplot(data=d1.hex, x=G, y=SO, geom="point", size=Freq) + 
  scale_area("Frequency", range=c(1, 4), trans="log", breaks=c(1, 10, 100, 1000, 10000))+ theme_bw() + xlab("Games (Count)")+ ylab("Strike Outs (Count)") + theme(legend.position="bottom")
  ggsave("./figure/BinScatter.pdf", width=5, height=5)

@@

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE),
        data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]),                                                                                    
        l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
#        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5, height=4, units="in")
@

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Bin Attribute Diagram]{\includegraphics[keepaspectratio=TRUE, width=4in]
  {./figure/Binning-DataVisualNumericalCenter.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Loss in Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE, eval=FALSE>>=
# bins <- rbind(c(1,1), c(2,2), c(2,5), c(5, 2), c(4,4), c(3, 5), c(5, 3), c(5, 5), c(5, 25), c(25, 5), c(2, 50), c(50, 2), c(5, 50))
computeLoss <- function(binning, data){
  lRandom <- lossCalc(data, binning=binning, type="random", newData = TRUE)
  lMean <- lossCalc(data, binning=binning, type="standard", newData=TRUE)
  datplot <- rbind(cbind(lRandom$NewData, Algorithm="Random"), cbind(lMean$NewData, Algorithm="Standard"))
    qplot(G, SO, fill = Freq, geom="tile", data = datplot, 
        main=paste("Baseball: Binning by (", binning[1], ", ", binning[2],")", sep="")) + facet_grid(.~Algorithm) +
        scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 5, 50, 500, 5000)) + theme_bw() + theme(legend.position="bottom")
    ggsave(paste("./figure/RandomStd", binning[1], binning[2], ".pdf", sep=""), width=8, height=5, units="in", dpi=1200)
  return(rbind(
    cbind(binX = binning[1], binY = binning[2], Function = "Random", lRandom$Loss), 
    cbind(binX = binning[1], binY = binning[2], Function = "Standard", lMean$Loss)))
}

@@

\textbf{Loss table for baseball example moved from beginning, this should not be its final resting place}
\noindent
%Table \ref{tablelossexample} gives an overview of the size of all different sources of loss for the examples in figure~\ref{binning}.

<<tablelossexample, echo=FALSE, eval=FALSE, results='asis',cache=FALSE>>=
loss0.1 <- lossCalc(data=X[,c(1:2)], binning=c(.1, .1), type="standard", newData=TRUE)
loss0.25 <- lossCalc(data=X[,c(1:2)], binning=c(.25, .25), type="standard", newData=TRUE)
loss0.5 <- lossCalc(data=X[,c(1:2)], binning=c(.5, .5), type="standard", newData=TRUE)
loss1 <- lossCalc(data=X[,c(1:2)], binning=c(1, 1), type="standard",newData=TRUE)

@@

table.loss <- data.frame(rbind(c(binwidth=0, TotalLoss.x=0, TotalLoss.y=0, TotalLoss.LogFreq=0, rows=200), 
                               c(binwidth=.1, loss0.1[5:7]*100, loss0.1[8]), 
                               c(binwidth=.25, loss0.25[5:7]*100, loss0.25[8]), 
                               c(binwidth=.5, loss0.5[5:7]*100, loss0.5[8]), 
                               c(binwidth=1, loss1[5:7]*100, loss1[8])))
names(table.loss) <- c("Bin Width", "% Loss in X", "% Loss in Y", "% Loss in log(Freq)", "Size of Data")
print(xtable(table.loss, digits=c(0, 2, 4, 4, 4, 0), caption="Loss due to binning for all examples in figure 4.", label="tablelossexample"), include.rownames=FALSE, table.placement="hbtp")
@
